---
title: "Public Health and Economic Impact of Natural Disasters in the United States"
author: "Brett A. Ochs"
date: "January 30, 2016"
output: html_document
---

# Synopsis

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

# Data Processing

## Download and Import Raw Data
```{r}
## Set variables to download and rename .csv files 
f.Name <- "repdata-data-StormData.csv.bz2"
f.URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
## Check to see if .csv file is in working directory 
## If no file exists the zipped .csv will be downloaded
## Data frame is loaded with .csv containing raw data
# if (file.exists(f.Name)) {
#     df <- read.csv(bzfile(f.Name))
# } else{
#     download.file(f.URL, f.Name)
#     df <- read.csv(bzfile(f.Name))
# }
```

## Data Processing Steps

# Results

```{r, echo=FALSE}
```

# Conclusion
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
